---
title: 'Intro to *data wrangling*' 
subtitle: 'dplyr & tidyr workshop'
author: "Samuel Robinson, Ph.D."
date: "November 5, 2020"
output: 
  beamer_presentation:
    theme: "default"
    colortheme: "lily"
    highlight: "tango"
urlcolor: blue
df_print: kable
header-includes: 
  \definecolor{darkturquoise}{rgb}{0.0, 0.81, 0.82}
  \useinnertheme{circles}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE)
library(tidyverse)
theme_set(theme_classic())
```

Normal data manipulation in R
---

\tiny

```{r, echo=T, eval=F}
#Changes species to factor
plants$Species <- as.factor(plants$Species) 
#Changes plant code to factor
plants$Plant.Code <- as.factor(plants$Plant.Code) 
#Changes Seed to factor
seeds$Seed <- as.factor(seeds$Seed) 
#Changes plant code to factor
seeds$Plant.Code <- as.factor(seeds$Plant.Code)
#Selects Flower, Code, Total.Germ columns
germ <- germ[,c('Flower','Code','Total.Germ')] 

#Sets numerics
plants[,c(3:9)] <- as.numeric(unlist(plants[,c(3:9)])) 
#Sets Dates
seeds$Collection.Date <- as.Date(seeds$Collection.Date,origin='2012-01-01')
```

Data manipulation using dplyr/tidyr
---

\tiny

```{r, echo=T,eval=F}
library(tidyverse)
#Convert factors in plants df
plants <- mutate(plants,Species=factor(Species),Plant.Code=factor(Plant.Code))
#Convert factors in seeds df
seeds <- mutate(seeds,Seed=factor(Seed),Plant.Code=factor(Plant.Code))
#Select Flower, Code, and Total.Germ columns in germ df
germ <- germ %>% select(Flower,Code,Total.Germ)

#Change columns 3:9 to numeric
plants <- plants %>% 
  mutate_at(vars(3:9),funs(as.numeric))
#Convert Collection.Date to Date format
seeds <- mutate(Collection.Date=as.Date(Collection.Date,origin='2012-01-01'))
```

\normalsize

- More compact, less typing
- Easier to read
- Faster (matters for large datasets)

How does this work?
---
Things to learn today:

- Basic syntax and table verbs
- Piping
- Reshaping
- Grouping
- Exercise!
- Final remarks

Basic Syntax
---
Both dplyr and tidyr work with data frames or tibbles

- data frame: similar to matrix, but with different data types for each column
- tibble: "compact" data frame, with some annoying features removed

\tiny
```{r,echo=T,eval=T}
head(iris) #Regular data frame
```

Basic Syntax
---
\tiny
```{r,echo=T,eval=T,message=F}
as_tibble(iris) #This is usually done automatically
```
Basic verbs - subsetting
---
- **select**: returns only columns that you want

\tiny

```{r,echo=T,eval=T,message=F}
head(iris,3) 
#Select Petal.Length,Petal.Width,Species
irisTemp <- select(iris,Petal.Length,Petal.Width,Species)
head(irisTemp,3)
```

Basic verbs - subsetting
---
Helper functions for **select**: *colon* operator

\tiny

```{r,echo=T,eval=T,message=F}
irisTemp <- select(iris,Petal.Length:Species) 
head(irisTemp,3) #All columns between Petal.Length and Species
irisTemp2 <- select(iris,Petal.Length,Petal.Width,Species) 
head(irisTemp2,3) #This is the same thing
```

Basic verbs - subsetting
---
Helper functions for **select**: -, and *contains*

\tiny

```{r,echo=T,eval=T,message=F}
irisTemp <- select(iris,-Species) 
head(irisTemp,3) #Selects all columns EXCEPT Species
irisTemp2 <- select(iris,contains('Petal'))
head(irisTemp2,3) #Selects columns with names containing 'Petal'
```

Basic verbs - subsetting
---
- **filter**: returns only rows that you want

\tiny

```{r,echo=T,eval=T,message=F}
head(iris,3) 
irisTemp <- filter(iris,Sepal.Length<5,Species=='versicolor')
head(irisTemp,3) #Chooses rows matching logical criteria
```

Basic verbs - make new variables
---
- **mutate**: add variables or alter existing ones

\tiny

```{r,echo=T,eval=T,message=F}
head(iris,3)
irisTemp <- mutate(iris,P.Width2=Petal.Width^2)
head(irisTemp,3) #Squares Petal.Width
```

Basic verbs - make new variables
---

\tiny

```{r,echo=T,eval=T,message=F}
head(irisTemp,3)
irisTemp <- mutate(iris,P.Width2=(Petal.Width^2)*2)
head(irisTemp,3) #Alters variable in place
```

Basic verbs - make new variables
---
- **mutate_at**: uses same function on columns of choice

\tiny

```{r,echo=T,eval=T,message=F,warning=F}
head(iris,3)
irisTemp <- mutate_at(iris,vars(Petal.Width,Petal.Length),funs(.^2))
head(irisTemp,3) #Squares Petal.Width & Length. "." means "data from column"
```

Basic verbs - make new variables
---

\tiny

- **rename** & **transmute**
```{r,echo=T,eval=T,message=F,warning=F}
irisTemp <- rename(iris,PWidth=Petal.Width,PLength=Petal.Length)
head(irisTemp,3) #Renames columns
irisTemp2 <- transmute(iris,P.Width2=(Petal.Width^2))
head(irisTemp2,3) #Same as mutate, but drops other columns
```

Exercises!
---
Using the *iris* dataset:

- Filter only "virginica" rows
- Make 2 new "area" columns, which are length $\times$ width of Petals and Sepals
- Get rid of all columns except "Species" + 2 new columns

\tiny

```{r,echo=F}
 iris %>% filter(Species=='virginica') %>%
  transmute(Species,P.Area=Petal.Length*Petal.Width,S.Area=Sepal.Length*Sepal.Width) %>% 
  head(10)
```


Piping - %>% 
---

\tiny

- Takes data from one verb and passes it to the next one
- Allows you to string together complex operations
```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
irisTemp <- select(iris,Sepal.Length,Species) %>% #Selects Sepal.Length & Species
  filter(Sepal.Length>5,Species=='versicolor') %>% #Filters using dataframe from above
  mutate(SLength2=Sepal.Length^2) #Mutates using dataframe from above
head(irisTemp)
```

Reshaping
---

\tiny


- This is very tedious to do in base R and Excel
- Reshaping operations in tidyr make this much easier
- Four main commands:
1. **gather** - gather columns into rows ('long format')
2. **spread** - spread rows into columns ('wide format')
3. **unite** - unite many columns into one (similar to *paste*)
4. **separate** - separates one column into many (similar to *strsplit*)

Reshaping - *gather*: columns to rows
---
- Make some data in "wide" format
```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
#Some fake data to work with
(fake <- data.frame(time = as.Date('2009-01-01') + 0:4,
                      X = rnorm(5, 0, 1),Y = rnorm(5, 0, 2)))
```

Reshaping - *gather*: columns to rows
---

- Change "wide" dataframe to "long" dataframe

\tiny

```{r,echo=T,eval=T,message=F,warning=F}
(fakeLong <- gather(fake,type, measurement, -time))
```

Reshaping - *spread*: rows to columns
---

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
fakeLong <- gather(fake,type, measurement, -time)
fakeLong %>% spread(type, measurement)
fakeLong %>% spread(time, measurement) #What has this done?
#Note: this must have unique row identifiers
```

Reshaping - *unite*: many columns into one
---

- Useful when combined with other reshaping functions

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
irisTemp <- iris %>% unite(newCol,Sepal.Length:Petal.Width,sep='_') 
head(irisTemp,10)

```

Reshaping - *separate*: one column into many
---

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
irisTemp %>% separate(newCol,c('SLength','SWidth','PLength','PWidth'),sep='_')
```

Reshaping - combinations of reshaping functions
---
Say we wanted lengths and widths in separate columns, split by Petal & Sepal

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
irisTemp <- iris %>% unite(sepals,Sepal.Length:Sepal.Width,sep='_') %>%
  unite(petals,Petal.Length:Petal.Width,sep='_') 
  head(irisTemp,10)
```

Reshaping - combinations of reshaping functions
---
- Now that measurements are *united*, we *gather* and then *separate* them

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
irisTemp %>% gather('Type','Measurement',sepals:petals) %>% 
  separate(Measurement,c('Length','Width'),sep='_',convert=T) %>% 
  head(10)
```

Exercises!
---
Using the *CO2* dataset:

- Select only _non-chilled_ plants from _Quebec_
- Pipe data frame to next command
- Change the uptake dataset from long to wide format (each plant should have its own column), with a column at the beginning showing concentration
- Hint: _filter_ rows and _select_ columns you need, then _spread_ to wide format

\tiny

```{r,echo=F}
CO2 %>% 
  filter(Type=='Quebec',Treatment=='nonchilled') %>% 
  select(-Type,-Treatment) %>% 
  spread(Plant,uptake)
```

Grouping 
---
- Often, we want to perform operations only on groups within data frames 
- For example, what is the average of each species' _Petal.width_?

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
with(iris,tapply(Petal.Width,Species,mean))
aggregate(Petal.Width~Species,data=iris,mean)
```

Grouping 
---
- How can this be done in dplyr/tidyr?

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
iris %>% group_by(Species) %>% 
  summarize(meanPWidth=mean(Petal.Width),sdPWidth=sd(Petal.Width))
```

\normalsize

- Apply *grouping*, then use *summary function*
- Data frame can be fed into other functions after summarizing

Grouping - Examples
---

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
iris %>% group_by(Species) %>% 
  summarize(count=n(),med=median(Petal.Width),iqr=IQR(Petal.Width))
```

\normalsize

- *n* is empty, because it uses the length of the subsetted data frame

Grouping - Examples
---

- Also useful for applying functions to subsets of data, *without* summarizing

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
iris %>% group_by(Species) %>% mutate(ID=1:n()) %>% #Makes ID column, with numbers 1-N
  filter(ID<4) #Selects ID 1-3 from each group
  
```

Grouping 
---

- Another way of doing the same thing

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F}
iris %>% group_by(Species) %>% 
  slice(1:3) #Selects rows 1-3 from each group
```

\normalsize

- You can use most of the subset and window functions across groups

Exercises!
---
Using the *InsectSprays* dataset:

- Find the mean and SD of each type of spray type
- Reshape dataframe so that each spray has its own column, with mean and SD in separate rows
- Hint: get summary stats first, then _gather_ and _spread_

\tiny

```{r,echo=F}
InsectSprays %>% group_by(spray) %>% 
  summarize(mean=mean(count),sd=sd(count)) %>% 
  gather('stat','meas',mean:sd) %>% 
  spread(spray,meas)
```

Final remarks {.columns-2 .smaller}
---

- dplyr & tidyr interface well with other parts of the tidyverse

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F,fig.height = 4, fig.width = 5, fig.align = "center"}
library(ggplot2)
CO2 %>% filter(conc==1000) %>% 
  group_by(Type,Treatment) %>% 
  summarize(meanUp=mean(uptake),
            maxUp=max(uptake),
            minUp=min(uptake)) %>% 
  #Code for ggplot begins here
  ggplot(aes(x=Type,col=Treatment))+
  geom_pointrange(aes(y=meanUp,ymax=maxUp,ymin=minUp))+
  labs(x='Area',y='Uptake at 1000ppm')+
  scale_colour_manual(values=c('red','blue'))
```

Final remarks {.columns-2 .smaller}
---
- dplyr & tidyr can pass data frames to and from other functions: use '.' operator

\tiny

```{r,echo=T,tidy=F,eval=T,message=F,warning=F,fig.height = 4, fig.width = 5, fig.align = "center"}
co2mod <- CO2 %>% 
  filter(Type=='Quebec') %>% data.frame() %>% 
  #Code for nls begins here
  nls(uptake~SSasymp(conc,A,B,C),
      start=list(A=40,B=-25,C=-5),data=.)

data.frame(conc=seq(50,1000,20)) %>% 
  predict(co2mod,newdata=.) %>% 
  data.frame(conc=seq(50,1000,20),predUp=.) %>% 
  #Code for ggplot begins here
  ggplot(aes(conc,predUp))+
  geom_line()+
  geom_point(data=filter(CO2,Type=='Quebec'),
             aes(conc,uptake))+
  labs(x='CO2 Concentration',y='Uptake')
  
```
