---
title: "Maximum likelihood and GLMs"
subtitle: "Well, how did I get here?"
author: "Samuel Robinson, Ph.D."
date: "November 26, 2020"
output: 
  beamer_presentation:
    theme: "default"
    colortheme: "lily"
    highlight: "tango"
df_print: kable
header-includes: 
  - \usepackage{tikz}
  - \definecolor{darkturquoise}{rgb}{0.0, 0.81, 0.82}
  - \useinnertheme{circles}
---

```{r setup, include=FALSE}
#Trick to get smaller R code size with out resorting to LaTeX text sizes
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

knitr::opts_chunk$set(echo = FALSE, eval = TRUE, message=TRUE, warning=TRUE, size = 'footnotesize')
library(tidyverse)
theme_set(theme_classic())
library(ggpubr)
library(knitr)
library(kableExtra)
library(latex2exp)

set.seed(123)

#Functions
logit <- function(x) log(x/(1-x))
invLogit <- function(x) exp(x)/(1+exp(x))

#Generate data that violate lm assumptions:
n <- 100
x <- runif(n,-10,10)
yhat <- 1 - 0.2*x #Expected value
y0 <- yhat + rnorm(n,0,2) #OK
y1 <- rpois(n,exp(yhat))  #Poisson process
y2 <- rbinom(n,1,invLogit(yhat))  #Binomial process

d1 <- data.frame(x,yhat,y0,y1,y2) #Dataframe


```

## Outline

- How to fit non-Bernoulli Binomial GLMs
- Residual checks - Deviance residuals
- Overdispersion
  - Negative binomial, Beta binomial
  - Zero-inflation - maybe a later lecture, along with occupancy models?
- Model selection
  - ML vs REML
  - log-likelihood and AIC

## R actually uses _log_ Likelihood

::: columns

:::: column

- Probabilities multiplied together quickly become _very small_
- Computers can't distinguish between extremely big or small numbers
- Therefore, it uses _log-likelihoods_ (also easier to calculate)

::::

:::: column

```{r, echo=FALSE, eval = TRUE, fig.width= 3, fig.height = 2}
ggplot() + geom_function(fun=llfun2) + xlim(-5,8) + labs(x=expression(paste('logit(',phi,')')),y=expression(paste('Likelihood(H,H,T|',phi,')'))) + geom_vline(xintercept=logit(2/3),linetype='dashed')

```

```{r, echo=FALSE, eval = TRUE, fig.width= 3, fig.height =2}

ggplot() + geom_function(fun=~log(llfun2(.))) + xlim(-5,8) + labs(x=expression(paste('logit(',phi,')')),y=expression(paste('log-likelihood(H,H,T|',phi,')'))) + geom_vline(xintercept=logit(2/3),linetype='dashed')

```

::::

:::