---
title: "Spatiotemporal models"
subtitle: '"Space is the place" - Sun Ra'
author: "Samuel Robinson, Ph.D."
date: "Oct 6, 2023"
output: 
  beamer_presentation:
    theme: "default"
    colortheme: "lily"
    highlight: "tango"
urlcolor: blue
df_print: kable
classoption: aspectratio=169   
header-includes: 
  - \definecolor{darkturquoise}{rgb}{0.0, 0.81, 0.82}
  - \useinnertheme{circles}
  - \let\oldShaded\Shaded %Change fontsize of R code chunks
  - \let\endoldShaded\endShaded
  - \renewenvironment{Shaded}{\scriptsize\oldShaded}{\endoldShaded}
  - \let\oldverbatim\verbatim %Change fontsize of code chunk output
  - \let\endoldverbatim\endverbatim
  - \renewenvironment{verbatim}{\tiny\oldverbatim}{\endoldverbatim}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 3: ST RE (move to other lecture)

## Spatial and Temporal Random Effects

"_My data were sampled over time or over a geographic area (or both). Can I just use day or site as a random effect?_"

- Short answer: "Yes"
- Long answer: You might be able to do better, because of the __1st Law of Geography__:

_"...everything is related to everything else, but near things are more related than distant things."_ Waldo Tobler

- If you have spatial or temporal information, this can help R to estimate random effects more accurately
  - Can improve prediction accuracy (smaller p-values)
  - Can give you hints about the underlying causal mechanisms

## Temporal or Spatial Data

- Correlation is often present in temporal data or spatial data; causes may be unknown or "uninteresting"
- Usually we are interested in accounting for these patterns, in order to better estimate the "interesting" patterns on top of them

::: columns

:::: column

```{r, fig.height=4, fig.width=4}
# data.frame(year=as.numeric(time(lynx)),lynxNum=as.numeric(lynx)) %>%
#   ggplot(aes(x=year,y=lynxNum))+geom_line()+
#   labs(x='Year',y='Lynx Numbers')
plot(sunspots,ylab='Sunspots per year')
```

::::

:::: column

```{r, fig.height=4, fig.width=4}
# ggplot(seals,aes(x=long,y=lat,fill=delta_long))+
#   geom_raster()+
#   scale_fill_continuous(type='viridis')+
#   labs(x='Longitude',y='Latitude')+
#   theme(legend.position = c(0.2,0.7))
image(volcano)
```

::::

:::

## Covariance

- Normal distributions\footnotemark don't just have a single $\sigma$, but a matrix of values
- If our data _y_ are _independent_, then it looks like this:

\begin{equation*}
y \sim Normal(\textcolor{orange}{M},\textcolor{red}{\Sigma})
\end{equation*}

\begin{equation*}
\textcolor{orange}{M} = [\mu_1, \mu_2, \mu_3]
\end{equation*}

\begin{equation*}
\textcolor{red}{\Sigma} = \begin{bmatrix}
\textcolor{red}{\sigma}^2 & 0 & 0 \\
0 & \textcolor{red}{\sigma}^2 & 0 \\
0 & 0 & \textcolor{red}{\sigma}^2
\end{bmatrix}
\end{equation*}

- Zeros mean "$\mu_1$, $\mu_2$, \& $\mu_3$ aren't related to each other"
- Diagonal elements = _variance_, off-diagonal = _covariance_

\footnotetext{Multivariate Normal}

## Covariance and Correlation

In real life, things may not be independent from each other. For example:

- $\textcolor{red}{\sigma}$ = 2 (variance = $\textcolor{red}{\sigma}^2$ = 4)
- $\mu_1$ and $\mu_2$ are strongly correlated (r=0.7), but $\mu_3$ is not related to anything (r=0). Shown here as a _correlation matrix_ ($\textcolor{red}{R}$):

\begin{equation*}
\textcolor{red}{R} = \begin{bmatrix}
1 & 0.7 & 0 \\
0.7 & 1 & 0 \\
0 & 0 & 1
\end{bmatrix}
\end{equation*}

- When multiplied by the variance, this becomes the _covariance matrix_ ($\textcolor{red}{\Sigma}$)

\begin{equation*}
\textcolor{red}{\Sigma} = \begin{bmatrix}
\textcolor{red}{\sigma}^2\times1 & \textcolor{red}{\sigma}^2\times0.7 & \textcolor{red}{\sigma}^2\times0 \\
\textcolor{red}{\sigma}^2\times0.7 & \textcolor{red}{\sigma}^2\times1 & \textcolor{red}{\sigma}^2\times0 \\
\textcolor{red}{\sigma}^2\times0 & \textcolor{red}{\sigma}^2\times0 & \textcolor{red}{\sigma}^2\times1
\end{bmatrix} = \begin{bmatrix}
4 & 2.8 & 0 \\
2.8 & 4 & 0 \\
0 & 0 & 4
\end{bmatrix}
\end{equation*}




## Gaussian Process Modelling

::: columns

:::: column

- We can model covariance between things as a function of _distance_, either in time or space
- Squared-exponential is fairly common\footnotemark:

\begin{equation*}
\begin{split}
\textcolor{red}{\Sigma} = & covariance \\
\textcolor{red}{\Sigma} = & variance \times correlation \\
\textcolor{red}{\Sigma} = & \textcolor{red}{\sigma}^2 \times e^{-\textcolor{green}{\rho}^2 Dist^2}\\
\end{split}
\end{equation*}

- Instead of finding a single $\textcolor{red}{\sigma}$ value, R now looks for $\textcolor{red}{\sigma}$ (maximum covariance) and $\textcolor{green}{\rho}$ (decay with distance)

::::

:::: column

```{r, fig.height=5,fig.width=4}
expand.grid(dist=seq(0,5,0.1),rho=c(0.5,1,2),sigma=c(1,2)) %>%
  mutate(cVal=(sigma^2)*exp((-rho^2)*(dist^2))) %>%
  mutate(category=paste0('sigma=',sigma,', rho=',rho)) %>%
  mutate(rho=factor(rho)) %>%
  ggplot(aes(x=dist,y=cVal,group=category,col=rho))+geom_line(size=1)+
  labs(x='Distance (km)',y='Covariance')+scale_colour_manual(values=c('red','purple','blue'))+
  theme(legend.position = c(0.8,0.8))
```

::::

:::

\footnotetext{Also common: AR-1 (temporal processes), Mat\'{e}rn (spatial processes)}

## Spatial random effects

::: columns

:::: column

- Say that we collected data at 16 sites, and we're interested in the effect of _y_ on _x_
- Let's first fit a model with a random intercept for site

```{r, echo=TRUE,size='tiny'}
#Same syntax as lmer models:
lmm2 <- glmmTMB(y~x+(1|site),data=dat2)
```

::::

:::: column

- If we plot the intercepts for each site, we see that they are clustered:

```{r,echo=FALSE,fig.height=3,fig.width=4,size='tiny'}
dat2 %>% select(site,lat,lon) %>% distinct() %>% arrange(site) %>% mutate(int=lmm2$sdr$par.random) %>%
  ggplot(aes(lon,lat,col=int))+geom_point(aes(size=abs(int)))+
  labs(title='Random intercepts (1|Site)',x='Longitude',y='Latitude',col='Intercept')+
  scale_colour_gradient2(low='blue',mid='purple',high='red')+
  guides(size=FALSE)
```

::::

:::


## Spatial random effects (cont.)

::: columns

:::: column

- Re-fit model with a spatial (exponential) random effect

```{r, echo=TRUE,size='tiny'}
#Coordinates
dat2$coords <- numFactor(dat2$lon,dat2$lat)

#Group factor (only 1 here)
dat2$group <- factor(rep(1,nrow(dat2)))

#Fit model with spatial random effect
lmm3 <- glmmTMB(y~x+exp(coords+0|group),data=dat2)
```

::::

:::: column

- Clustering effect modeled as a spatial random effect

```{r, echo=FALSE, fig.width=4,fig.height=3,size='tiny'}
#Plot spatial random effect
spRanEff <- expand.grid(lon=-10:10,lat=-10:10) %>%
  mutate(coords=numFactor(lon,lat),group=factor(rep(1,nrow(.))),x=0) %>%
  mutate(pred=predict(object=lmm3,newdata=.,type='response',allow.new.levels = TRUE))

ggplot(spRanEff,aes(lon,lat,fill=pred))+geom_raster()+
  geom_point(data=dat2,aes(lon,lat,fill=NULL),col='red')+
  labs(x='Longitude',y='Latitude',title='Spatial random effect',fill='Intercept')

```


::::

:::


## A challenger approaches

- Ho ho ho! Merry Christmas! In order to maximize the number of presents that you get from Santa Claus, you've decided to apply an analytic approach, and have collected data across Alberta on _number of Christmas presents received_
- You've also collected data on things that might influence Saint Nick's generosity (_naughtiness_, _presence of milk and cookies_, _chimney width_)
- Fit a GLMM to the present data, one using spatial random intercepts, and one using "regular" random intercepts
- Which type of snack should you leave out for Santa? Which area might you consider moving to??

