---
title: "ggplot2"
subtitle: "_Much_ funnier if you speak Mandarin"
author: "Samuel Robinson, Ph.D."
date: "October 29, 2020"
output: 
  beamer_presentation:
    theme: "default"
    colortheme: "lily"
    highlight: "tango"
urlcolor: blue
df_print: kable
header-includes: 
  \definecolor{darkturquoise}{rgb}{0.0, 0.81, 0.82}
  \useinnertheme{circles}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggpubr)
# theme_set(theme_classic())
presTheme <- theme(axis.title=element_text(size=30),axis.text=element_text(size=10))
theme_set(presTheme)

set.seed(123)
```

## Motivation 

What is `ggplot2`?

- `ggplot` philosophy
- Simple plots
- Some useful techniques
- More complicated plots
  
## What is `ggplot2`?

- Updated version of `ggplot` (older `R` package)
- Implementation of Wilkinson's _grammar of graphics_
- Elements: data, transformations, elements, scale, guide, coordinates
- Describes a [layered approach to building graphics](https://doi.org/10.1198/jcgs.2009.07098) beyond formulaic plots (e.g. "boxplot", "scatterplot")
- Many different extensions available [here](https://exts.ggplot2.tidyverse.org/gallery/)
  
Philosophy:

- Data input centered around around `data.frames`
- Data display centered around `geoms` (geometric objects)
- Columns from data frames are `mapped` into geoms using `aesthetics`
- geoms are displayed according to `themes`

## Simple example - scatterplot

\tiny 

```{r, echo=TRUE} 
data(mtcars) # mtcars dataset (built into R)
head(mtcars,5) # Show first 5 rows
```

::: columns

:::: column

```{r, eval=FALSE, echo = TRUE}
# Load ggplot library
library(ggplot2) 

# Top line of code says: 
# - data from mtcars dataframe
# - aes = aesthetics from dataframe
# - map disp to x-axis, mpg to y-axis
ggplot(data = mtcars, aes(x = disp, y = mpg))+
  geom_point() # Display data using points
```

::::

:::: column

```{r echo=FALSE}
ggplot(data=mtcars,aes(x=disp,y=mpg))+geom_point()
```

::::

:::

## Simple example - bar plot

\tiny 

```{r, echo=TRUE} 
data(mtcars) # mtcars dataset (built into R)
head(mtcars,5) # Show first 5 rows
```

::: columns

:::: column

```{r, eval=FALSE, echo = TRUE}
# Top line of code says: 
# - map gear (number of gears) to x-axis
#   - first converted to a factor
ggplot(data = mtcars, aes(x = factor(gear)))+ 
  geom_bar() 
# Display number of data points for each factor

# Automatically uses stat='count' to group
#   data according to factor
```

::::

:::: column

```{r echo=FALSE}
ggplot(data=mtcars,aes(x=factor(gear)))+ 
  geom_bar()
```

::::

:::

## Simple example - histogram

\tiny 

```{r, echo=TRUE} 
data(mtcars) # mtcars dataset (built into R)
head(mtcars,5) # Show first 5 rows
```

::: columns

:::: column

```{r eval=FALSE, echo=TRUE}
# Top line of code says: 
# - map disp (displacement) to x-axis
ggplot(data = mtcars, aes(x = disp))+ 
  # Group disp into bins, and display 
  #   count in each bin
  geom_histogram() 
```

::::

:::: column

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=mtcars,aes(x=disp))+ geom_histogram()
```

::::

:::

## Simple example - histograms and density plots

\tiny

::: columns

:::: column

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Histogram
ggplot(data=mtcars,aes(x=disp))+ 
  geom_histogram() 
```

\normalsize

Histogram

::::

:::: column

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Density plot
ggplot(data=mtcars,aes(x=disp))+ 
  geom_density() 
```

\normalsize

Probability density plot

$a\int_{-\infty}^{\infty} f(x) \,dx=1$

::::

:::

## Colours in plots

- Colours can be _mapped_ (via `aes`) or _set_ (outside of `aes`)
 
\tiny 

::: columns

:::: column

```{r, echo = TRUE}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  # Maps gear to colour
  geom_point(aes(col=factor(gear))) 
```

::::

:::: column

```{r echo=TRUE}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  geom_point(colour='red') #Sets colour
```

::::

:::

\normalsize

- Notice how `aes` was used twice in Figure 1? If used within the `ggplot` command, the rest of the geoms will remember it. Used within a geom, it will _update_ the aesthetic

## What if I want different colours?

::: columns

:::: column

- Default colour themes are pretty bad. Change them with `scale_colour_manual` 
- Use `scale_fill_manual` for area-based colours (e.g. bar plots, polygons)
- Remember, 10% of males are red-green colourblind!

::::

:::: column

```{r echo=FALSE, eval = TRUE, fig.width=5, fig.height=5}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  geom_point(aes(col=factor(gear)))+
  scale_colour_manual(values=c('blue','purple','red'))
```

::::

:::

\small

```{r, echo = TRUE, eval = FALSE}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  geom_point(aes(col=factor(gear)))+
  scale_colour_manual(values=c('blue','purple','red'))
```


## Sizes in plots

- Sizes can also be _mapped_ (via `aes`) or _set_ (outside of `aes`)
 
\tiny 

::: columns

:::: column

```{r, echo = TRUE}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  # Maps gear to size
  geom_point(aes(size=gear)) 
```

::::

:::: column

```{r echo=TRUE}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  geom_point(size=10) #Sets size
```

::::

:::

\normalsize

- Similar to colour choices, you can alter mapped sizes using `scale_size`

## Change plot `theme`

::: columns

:::: column

- `theme` controls almost all non-data elements of plots
- Made up of _elements_: `element_line()`, `element_text()`, `element_rect()`
- Let's make some changes:

\tiny

```{r, echo = TRUE, eval=FALSE}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  # Maps gear to colour
  geom_point(aes(col=factor(gear))) +
  #Changes plot theme
  theme(axis.title.x=element_text(size=10), 
    legend.background=element_rect(fill='red'),
    legend.position=c(0.8,0.8),
    panel.grid.minor=element_line(colour='purple'),
    panel.grid.major=element_line(colour='orange'))
```

::::

:::: column

```{r, echo = FALSE, eval=TRUE, fig.width=4,fig.height=5}
ggplot(data=mtcars,aes(x=disp,y=mpg))+
  # Maps gear to colour
  geom_point(aes(col=factor(gear))) +
  #Changes plot theme
  theme(axis.title.x=element_text(size=10), 
    legend.background=element_rect(fill='red'),
    legend.position=c(0.8,0.8),
    panel.grid.minor=element_line(colour='purple'),
    panel.grid.major=element_line(colour='orange'))
```

\small 

- This plot is hideous, but it gives you the idea!
- Use `?theme` to see all options

::::

:::

## Preset themes

```{r, echo = FALSE, eval=TRUE, out.width='100%'}
p <- ggplot(data=mtcars,aes(x=disp,y=mpg)) + geom_point() 
p1 <- p + theme_grey() + annotate('text',y=30,x=400,label='theme_grey()') #Regular grey theme
p2 <- p + theme_bw() + annotate('text',y=30,x=400,label='theme_bw()') #B/W
p3 <- p + theme_classic() + annotate('text',y=30,x=400,label='theme_classic()') #Classic
p4 <- p + theme_minimal() + annotate('text',y=30,x=400,label='theme_minimal()') #Minimalistic
ggarrange(p1,p2,p3,p4,ncol=2,nrow=2)
```

## Make your own themes!

::: columns

:::: column

- You can modify existing themes in order to create your own
- Try using `theme_set()` at the start of your script to pre-set the theme for the rest of the script

\tiny

```{r, echo = TRUE, eval=FALSE}
myTheme <- theme_classic()+ #Existing theme
  #Makes axis text bigger
  theme(axis.title=element_text(size=30),
        axis.text=element_text(size=10),
        legend.position='bottom')
#Sets up this theme as "default"
theme_set(myTheme)
```

::::

:::: column

```{r, echo = FALSE, eval=TRUE, fig.width=4, fig.height=4}
myTheme <- theme_classic()+ #Existing theme
  #Makes axis text bigger
  theme(axis.title=element_text(size=30),
        axis.text=element_text(size=10),
        legend.position='bottom')
  

ggplot(data=mtcars,aes(x=disp,y=mpg))+
  # Maps cyl to colour
  geom_point(aes(col=factor(cyl))) +
  scale_colour_manual(values=c('blue','purple','red')) +
  myTheme

theme_set(theme_grey())

```

::::

:::

## Complex plots - facets

\small

- It is possible to break up the plot into smaller `facets` that are mapped to a given variable
- This can be combined with colour/size mappings

\tiny
```{r, echo = TRUE, eval = TRUE, fig.width=6, fig.height=3.5}
ggplot(mtcars,aes(x=disp,y=mpg))+ geom_point(aes(col=wt))+
  facet_grid(factor(am) ~ factor(gear))
```

## Complex plots - summary statistics (mean)

\tiny
```{r, echo = TRUE,fig.width=8,fig.height=5}
ggplot(mtcars,aes(x=factor(am),y=mpg))+ 
  geom_point(position=position_jitter(width=0.05))+ #Adds noise to data in x-dimension
  geom_point(stat='summary',fun=mean,col='red',size=3) #Mean only
```

## Complex plots - summary statistics (mean + SD)

\tiny
```{r, echo = TRUE,fig.width=8,fig.height=5}
ggplot(arrange(mtcars,am,disp),aes(x=factor(am),y=mpg))+ 
  geom_point(position=position_jitter(width=0.05))+
  geom_pointrange(stat='summary',fun.data=mean_se,
                  fun.args = list(mult = 2),col='red') #Mean + 2 SE
```

## Complex plots - smoothers

- You can add `lm` (or other model) predictions to your plots:

::: columns

:::: column
\tiny 

```{r, echo = TRUE, eval = FALSE}
ggplot(mtcars,aes(x=disp,y=mpg))+ 
  geom_point() + 
  geom_smooth(method='lm',formula=y~x)
```

```{r, echo = FALSE, eval = TRUE, fig.width=6,fig.height=6}
ggplot(mtcars,aes(x=disp,y=mpg))+ 
  annotate('text',y=30,x=400,label='linear model')+
  geom_point() + geom_smooth(method='lm',formula=y~x)
```

::::

:::: column

\tiny 

```{r, echo = TRUE, eval = FALSE}
ggplot(mtcars,aes(x=disp,y=mpg))+ 
  geom_point() + 
  geom_smooth(method='gam',formula=y~s(x))
```
``` {r, echo = FALSE, eval = TRUE,fig.width=6,fig.height=6}
ggplot(mtcars,aes(x=disp,y=mpg))+ 
  annotate('text',y=30,x=400,label='GAM smoother')+
  geom_point() + geom_smooth(method='gam',formula=y~s(x))
```

:::: 

:::

## Complex plots - transformations

- You can show transformed data OR you can transform the axes themselves using `scale_*_log10` (x or y axis)

::: columns

:::: column
\tiny 

```{r echo = TRUE, eval = FALSE}
ggplot(mtcars,aes(x=log(disp),y=log(mpg)))+ 
  geom_point() + 
  geom_smooth(method='lm',formula=y~x)
# Harder to interpret, because people can't
#     usually do log(x) in their head
```

```{r, echo = FALSE, eval = TRUE, fig.width=6,fig.height=6}
ggplot(mtcars,aes(x=log(disp),y=log(mpg)))+ 
  geom_point() + 
  geom_smooth(method='lm',formula=y~x)
```

::::

:::: column

\tiny 

```{r, echo = TRUE, eval = FALSE}
ggplot(mtcars,aes(x=disp,y=mpg))+ 
  geom_point() + 
  geom_smooth(method='lm',formula=y~x)+
  scale_x_log10() + scale_y_log10() 
# sqrt is also popular
```

```{r, echo = FALSE, eval = TRUE,fig.width=6,fig.height=6}
ggplot(mtcars,aes(x=disp,y=mpg))+ 
  geom_point() + 
  geom_smooth(method='lm',formula=y~x)+
  scale_x_log10() + scale_y_log10() # sqrt is also popular
```

:::: 

:::

## Things to remember:

- Simpler plots are often better. Try to keep it to 3 aesthetics per panel. Avoid 3D plots.
- Making plots is iterative. Make a simple one and tweak it to improve it.
- Avoid "non-data ink" (see [Edward Tufte's](https://www.amazon.ca/Visual-Display-Quantitative-Information/dp/1930824130) work)
- Our eyes are good at estimating linear positions, but bad at estimating area, volume, colour shading, and angles:

```{r, echo = FALSE, eval = TRUE,fig.width=12,fig.height=4}
d1 <- data.frame(x=factor(c('a','b','c')),y=c(23,46,31))

p1 <- ggplot(d1,aes(x=x,y=y))+geom_col()+theme_classic()
p2 <- ggplot(d1,aes(x=x,y=1))+geom_point(aes(size=y))+theme_classic()+scale_size(range=c(5,15))
p3 <- ggplot(d1, aes(x="", y=y, fill=x)) + geom_bar(width = 1, stat = "identity")+coord_polar("y") + theme_void() + theme(axis.text.x=element_text(size=0))

ggarrange(p1,p2,p3,ncol=3)

```

## A challenger approaches:

Make these figures! Datasets are found in `mpg`, `msleep`, `trees`, and `starwars` (built into the `ggplot2` and `dplyr` packages)

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%'}
p1 <- ggplot(mpg,aes(x=displ,y=cty))+ 
  geom_point() + geom_smooth(method='gam',formula=y~s(x),col='red')+
  theme_classic()

p2 <- ggplot(msleep,aes(x=bodywt,y=brainwt))+
  geom_point(aes(col=vore))+
  scale_x_log10()+scale_y_log10()+geom_smooth(method='lm',formula=y~x,col='black')+
  labs(x='Body weight (kg)',y='Brain Weight (kg)',col='Feeding')+
  theme_bw()

p3 <- ggplot(trees,aes(x=Girth,y=Height,col=Volume,size=Volume))+
  geom_point()+theme_grey()

p4 <- ggplot(starwars,aes(x=height))+geom_histogram()+facet_wrap(~sex)+ theme_minimal()+labs(x='Height of Star Wars Characters (cm)')

ggarrange(p1,p2,p3,p4,ncol=2,nrow=2)

```